Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-28 paper.tex
Update development model.

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\tskit}[0]{\texttt{tskit}}
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\scrm}[0]{\texttt{scrm}}
\newcommand{\SLiM}[0]{\texttt{SLiM}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\SeqGen}[0]{\texttt{Seq-Gen}}
\newcommand{\Pyvolve}[0]{\texttt{Pyvolve}}
\newcommand{\phastsim}[0]{\texttt{phastsim}}

@@

Currently there are a small number of other simulators that will generate
coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
Cosi2~\citep{shlyakhter2014cosi2}, and discoal~\citep{kern2016discoal}.
Each of these packages has their own benefits and possible drawbacks,
however none of them at present will output tree sequence formatted data.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/sweeps-perf}
\end{center}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \texttt{discoal} for simulating selective sweeps.
}
\end{figure}

We have implemented selective sweeps in \msprime\ similarly to
the \texttt{discoal} implementation-- sweep trajectories are
generated from a jump process approximation to the conditional diffusion
of an allele bound for fixation~\citep{coop2004ancestral}.
Details of our implementation are available in the Appendix.
This implementation allows us to further generalize sweep models to
changing population sizes as well as selection from standing variation,
both of which are features that are planned for the near future.
Moreover because this implements the basic infrastructure for the structured
coalescent, adding support for, e.g.,
inversions is straightforward~\citep{peischl2013sequential}.

@@

We support changing the simulation model at specified times so that we can
easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}. We
can combine any number of different simulation models, allowing for the
flexible choice of simulation scenarios without severely damaging efficiency.
As the discrete time Wright-Fisher model improves accuracy of genealogical
patterns in the recent past, we can simulate the recent history using this
model and then switch to the standard coalescent to more efficiently simulate
the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the perfomance of
this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
hybrid approach provides large efficiency gains.

\subsection*{Development model}
The community development process for \msprime\ is enabled by following an
open source development practise with a strong emphasis on code quality
and correctness. To ensure a consistent style, we require that all code
follows the PEP8 guidelines. Unit tests for all additions are required, and
are automatically run on each pull request on a variety of continuous
integration testing platforms. For new simulation features, statistical tests
must also be added, comparing the distribution of results to either analytical
results or existing simulators.

>>>>

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\tskit}[0]{\texttt{tskit}}
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\scrm}[0]{\texttt{scrm}}
\newcommand{\SLiM}[0]{\texttt{SLiM}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\SeqGen}[0]{\texttt{Seq-Gen}}
\newcommand{\Pyvolve}[0]{\texttt{Pyvolve}}
\newcommand{\phastsim}[0]{\texttt{phastsim}}
\newcommand{\discoal}[0]{\texttt{discoal}}

@@

Currently there are a small number of other simulators that will generate
coalescent samples with sweeps including, SelSim~\citep{spencer2004selsim}
mbs~\citep{teshima2009mbs}, msms~\citep{ewing2010msms},
Cosi2~\citep{shlyakhter2014cosi2}, and \discoal~\citep{kern2016discoal}.
Each of these packages has their own benefits and possible drawbacks,
however none of them at present will output tree sequence formatted data.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/sweeps-perf}
\end{center}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \discoal\ for simulating selective sweeps.
}
\end{figure}

We have implemented selective sweeps in \msprime\ similarly to
the \discoal\ implementation-- sweep trajectories are
generated from a jump process approximation to the conditional diffusion
of an allele bound for fixation~\citep{coop2004ancestral}.
Details of our implementation are available in the Appendix.
This implementation allows us to further generalize sweep models to
changing population sizes as well as selection from standing variation,
both of which are features that are planned for the near future.
Moreover because this implements the basic infrastructure for the structured
coalescent, adding support for, e.g.,
inversions is straightforward~\citep{peischl2013sequential}.

@@

We support changing the simulation model at specified times so that we can
easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}. We
can combine any number of different simulation models, allowing for the
flexible choice of simulation scenarios without severely damaging efficiency.
As the discrete time Wright-Fisher model improves accuracy of genealogical
patterns in the recent past, we can simulate the recent history using this
model and then switch to the standard coalescent to more efficiently simulate
the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the performance of
this approach with ARGON~\citep{palamara2016argon}, a specialised DTWF
simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
hybrid approach provides large efficiency gains.

\subsection*{Development model}
\texttt{Msprime} has a large number of features, encompassing
the functionality of several more specialised simulators
while maintaining excellent performance.
It is developed by a geographically distributed team of volunteers under an
open source community development model, with a strong emphasis
on code quality and correctness. As in any large code base,
unit tests play a key role in ensuring that new additions behave
as expected and \msprime\ has an extensive suite.
As of the 1.0.0 release \msprime\ consists of around 13K lines of C and 11K
lines of Python, with suites of 122 C tests (7K lines of code) and
1350 Python tests (22K lines of code). These tests are run
automatically on different operating systems on each pull request
(where a contributor updates the codebase), using standard Continuous
Integration (CI) methodology. Other CI services check code formatting
and for common errors, as well as producing reports on the level of
test coverage, e.g., pinpointing areas where additional unit tests may
be required.

Unit tests are vital for ensuring software quality and correctness, but
they are usually of little value in assessing the statistical properties
of simulations. To validate the correctness of simulation output
we maintain a suite of statistical tests (as of 1.0.0,
217 validation tests, in 6K lines of code). These consist of running many
replicate simulations, and checking the properties of the output
against either analytical results of the output of other simulators
where such results are not available.
For example, simulations of complex demography are validated against \ms,
selective sweeps against \discoal\, and Wright-Fisher simulations
against forwards-time simulations in \SLiM.
This suite of tests is run before every release, to ensure
that statistical errors have not been introduced.
